---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SUFA

A `C++`-based implementation of the SUbspace Factor Analysis (SUFA) model by @chandra2023sufa.

<!-- badges: start -->

<!-- badges: end -->

## Installation

The package has been tried and tested in Ubuntu and macOS.

### Required UNIX Packages

-   `libgsl`

-   `openmp`

-   `R (>= 4.3.1)`

### Installation from Github

You can install the development version of SUFA from [GitHub](https://github.com/) with:

``` {r eval=F}
install.packages("devtools")
devtools::install_github("noirritchandra/SUFA", build_vignettes = TRUE)
```

Must set `build_vignettes = TRUE` to install the vignette files containing illustrations. 
However, this may take considerably more time to install the package.

## Simulation Example

Refer to the following vignette for illustrations in simulated examples:

```{r sim_example, eval=FALSE}
vignette(topic="Intro_simulation",package = "SUFA")
```


## Genedata Application

Refer to the following vignette in gene expressions datasets for inference on gene networks:

```{r genedata, eval=FALSE}
vignette(topic="Genedata_application",package = "SUFA")
```